---
title: "R Notebook - Analyse iLand output"
output: html_notebook
---

After having run the iLand simulations for the 12 clusters the resulting .sqlite files need to be evaluated. Goal is to see how the tree development of each cluster develops.

# 1. Load data from .sqlite files

To be able to load the data, fist a connection to the .sqlite file needs to be established. For this the library "RSQLite" is needed.

```{r librarys}
#install.packages("RSQLite")
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggpubr)
library(rstatix)
library(ggplot2)
```

With the necessary libraries loaded, the data can be assessed. First a connection to the database is established, afterwards the tables inside can be read.

```{r load_data}
setwd("rawData/iLand/iLand_output")
getwd()

# Get output dirs and run through them
lDirs <- list.dirs(recursive = FALSE)
lDirs <- grep("_new", lDirs, value = TRUE)
DBTables <- list()
DBT_sap <- list()
DBT_sapd <- list()
iFiles <- 1
iDirs <- 1
bExtSeed <- FALSE

# Outer loop runs through all directories at that path and loads .sqlite files
# into a list. Also a counter (iDirs) counts the number of directories accessed
for (l_dir in lDirs) {
  lFiles <- list.files(path = l_dir, pattern = ".sqlite", full.names = TRUE)
  
  # Inner loop runs through all files (.sqlite) in a directory and loads the
  # stand table into a variable.
  # File for cluster 3_2 is skipped, due to this stand being empty in run 
  # without external seeeds
  for (lFile in lFiles) {
    l_db <- dbConnect(RSQLite::SQLite(), lFile)
    l_dfStnd <- dbReadTable(l_db, "stand")
    l_dfSapl <- dbReadTable(l_db, "sapling")
    l_dfSapD <- dbReadTable(l_db, "saplingdetail")
    dbDisconnect(l_db)
    
    bExtSeed <- !grepl("noseed", lFile)

    if (nrow(l_dfStnd) == 0) {
      next
    }
    
    l_dfStnd$filenm <- tools::file_path_sans_ext(basename(lFile))
    l_dfStnd <- l_dfStnd %>%
      dplyr::select(filenm,
                    year,
                    species,
                    volume_m3,
                    dbh_avg_cm,
                    basal_area_m2,
                    count_ha,
                    cohort_count_ha,
                    cohort_basal_area)
    l_dfStnd$run_nr <- iDirs
    l_dfStnd$external_seed <- bExtSeed
    
    l_dfSapl$filenm <- tools::file_path_sans_ext(basename(lFile))
    l_dfSapD$filenm <- tools::file_path_sans_ext(basename(lFile))
    
    DBTables[[iFiles]] <- l_dfStnd
    DBT_sap[[iFiles]] <- l_dfSapl
    DBT_sapd[[iFiles]] <- l_dfSapD
    
    iFiles <- iFiles + 1
    #dbDisconnect(l_db)
  }
  iDirs <- iDirs + 1
}
# Bring all the data together in big_data variable
OG_big_data <- bind_rows(DBTables)
OG_sapl <- bind_rows(DBT_sap)
OG_sapd <- bind_rows(DBT_sapd)

# Cleanup of some variables
rm(DBTables, l_db, l_dfStnd, iDirs, iFiles, l_dir, lDirs, lFile, lFiles, 
   bExtSeed, DBT_sapd, DBT_sap, l_dfSapl, l_dfSapD)
```

# 2. Data handling

For further processing it might be usefull to split the filename into a few pieces:

-   Climate model

-   Cluster

-   With or without external seed

```{r data_handling}
# 1. stand table
big_data <- OG_big_data |>
  separate_wider_delim(filenm, delim = "_", names = c("clim_model",
                                                      "clim_scenario",
                                                      "env_clst",
                                                      "stnd_clst",
                                                      "ext_seed"))
big_data <- big_data %>% dplyr::select(!ext_seed)
big_data$cluster <- paste(big_data$env_clst, big_data$stnd_clst, sep = "_")
big_data <- big_data %>%
  dplyr::select(!c(env_clst, stnd_clst))
big_data$clim_scenario <- as.factor(big_data$clim_scenario)
big_data$clim_model <- as.factor(big_data$clim_model)
big_data$cluster <- as.factor(big_data$cluster)
big_data$species <- as.factor(big_data$species)

strCols <- c("volume_m3", "dbh_avg_cm", "basal_area_m2",
             "count_ha", "cohort_count_ha", "cohort_basal_area")

w_means <- big_data %>%
  dplyr::group_by(clim_scenario, external_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),.groups = "drop")

# 2. sapling table
big_sapl <- OG_sapl |>
  separate_wider_delim(filenm, delim = "_", names = c("clim_model",
                                                      "clim_scenario",
                                                      "env_clst",
                                                      "stnd_clst",
                                                      "ext_seed"))
big_sapl$cluster <- paste(big_sapl$env_clst, big_sapl$stnd_clst, sep = "_")
big_sapl$clim_model <- as.factor(big_sapl$clim_model)
big_sapl$clim_scenario <- as.factor(big_sapl$clim_scenario)
big_sapl$cluster <- as.factor(big_sapl$cluster)
big_sapl$species <- as.factor(big_sapl$species)
big_sapl$ext_seed <- as.factor(big_sapl$ext_seed)
big_sapl <- big_sapl %>% dplyr::select(!c(env_clst, stnd_clst, ru, rid))

strCols <- c("count_ha", "count_small_ha", "cohort_count_ha", "height_avg_m",
             "age_avg", "LAI")

w_means_sapl <- big_sapl %>%
  dplyr::group_by(clim_scenario, ext_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),.groups = "drop")

# 3. saplingdetails table
big_sapD <- OG_sapd |>
  separate_wider_delim(filenm, delim = "_", names = c("clim_model",
                                                      "clim_scenario",
                                                      "env_clst",
                                                      "stnd_clst",
                                                      "ext_seed"))
big_sapD$cluster <- paste(big_sapD$env_clst, big_sapD$stnd_clst, sep = "_")
big_sapD$clim_model <- as.factor(big_sapD$clim_model)
big_sapD$clim_scenario <- as.factor(big_sapD$clim_scenario)
big_sapD$cluster <- as.factor(big_sapD$cluster)
big_sapD$species <- as.factor(big_sapD$species)
big_sapD$ext_seed <- as.factor(big_sapD$ext_seed)
big_sapD <- big_sapD %>% dplyr::select(!c(env_clst, stnd_clst, ru, rid))

strCols <- c("age", "height", "dbh", "n_represented")

w_means_sapD <- big_sapD %>%
  dplyr::group_by(clim_scenario, ext_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),.groups = "drop")

# clean environment
rm(strCols)
```

## 2.1 Effect strength

Effect strength of all variables that influence the number of trees is calculated to identify which can be included in the mean calculation.

```{r effect_strength}
ggplot(w_means, aes(x = count_ha_mean)) + geom_histogram()
ggplot(w_means, aes(x = count_ha_mean)) + geom_density()

# test for varianzhomogeniocity (if p < 0.05 -> no homogeneity of variance)
w_means %>% levene_test(count_ha_mean ~ clim_scenario)  #homogeneity of variance
w_means %>% levene_test(volume_m3_mean ~ clim_scenario)  #homogeneity of variance
w_means %>% levene_test(count_ha_mean ~ external_seed)  #no homogeneity of var
w_means %>% levene_test(count_ha_mean ~ cluster)        #no homogeneity of var
w_means %>% levene_test(count_ha_mean ~ species)        #no homogeneity of var

# test for effect strength - var.equal = FALSE -> no homogeneity of variance
w_means %>% cohens_d(count_ha_mean ~ clim_scenario, var.equal = TRUE) #negligible effect
w_means %>% cohens_d(volume_m3_mean ~ clim_scenario, var.equal = TRUE) #negligible effect
w_means %>% cohens_d(count_ha_mean ~ external_seed, var.equal = FALSE) #small effect
w_means %>% cohens_d(count_ha_mean ~ cluster, var.equal = FALSE) #no to large effect
w_means %>% cohens_d(count_ha_mean ~ species, var.equal = FALSE) #no to large effect

ggplot(w_means[w_means$year == 30 & w_means$external_seed == FALSE,], 
       aes(x = species,y = count_ha_mean, fill = species)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = count_ha_mean - count_ha_sd,
                    ymax = count_ha_mean + count_ha_sd)) +
  facet_grid(cluster ~ clim_scenario, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Count per ha, in Year 30 (no external seed)")
```

Climate scenarios do not seem to have a huge influence on the dataset and can therefore calculated into the uncertainty.

## 2.2 Mean over all clim_scenarios

Calculate the mean values over all 4 climate scenarios.

```{r mean_clim}
# stand
strCols <- c("volume_m3", "dbh_avg_cm", "basal_area_m2",
             "count_ha", "cohort_count_ha", "cohort_basal_area")

w_means_clim <- big_data %>%
  dplyr::group_by(external_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),
                   .groups = "drop")

# saplings
strCols <- c("count_ha", "count_small_ha", "cohort_count_ha",
             "height_avg_m", "age_avg")

w_means_sapl_clim <- big_sapl %>%
  dplyr::group_by(ext_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),
                   .groups = "drop")

# sapling details
strCols <- c("n_represented", "dbh", "height", "age")

w_means_sapD_clim <- big_sapD %>%
  dplyr::group_by(ext_seed, cluster, year, species) %>%
  dplyr::summarise(across(strCols, list(mean = mean, sd = sd)),
                   .groups = "drop")

rm(strCols)
```

# 3. Plots

1.  Development volume, dbh, basal area, count and cohort count over the simulation extend.

```{r development_plots}
# volume/biomass
ggplot(w_means_clim, aes(x = year, y = volume_m3_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = volume_m3_mean - volume_m3_sd,
                  ymax = volume_m3_mean + volume_m3_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean volume(m³)/biomass per cluster")

# dbh
ggplot(w_means_clim, aes(x = year, y = dbh_avg_cm_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = dbh_avg_cm_mean - dbh_avg_cm_sd,
                  ymax = dbh_avg_cm_mean + dbh_avg_cm_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean dbh(cm) per cluster")

# basal area
ggplot(w_means_clim, aes(x = year, y = basal_area_m2_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = basal_area_m2_mean - basal_area_m2_sd,
                  ymax = basal_area_m2_mean + basal_area_m2_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean basal area (m²) per cluster")

# stem count
ggplot(w_means_clim, aes(x = year, y = count_ha_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = count_ha_mean - count_ha_sd,
                  ymax = count_ha_mean + count_ha_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean stem count (n/ha) per cluster")

# cohort count
ggplot(w_means_clim[w_means_clim$external_seed == FALSE,],
       aes(x = year, y = cohort_count_ha_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = cohort_count_ha_mean - cohort_count_ha_sd,
                  ymax = cohort_count_ha_mean + cohort_count_ha_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean cohort count (n/ha) per cluster (no seed)")

ggplot(w_means_clim[w_means_clim$external_seed == TRUE,],
       aes(x = year, y = cohort_count_ha_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = cohort_count_ha_mean - cohort_count_ha_sd,
                  ymax = cohort_count_ha_mean + cohort_count_ha_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean cohort count (n/ha) per cluster")
```

2.  Dominant species on each cluster in Year 30

```{r dominant_species_plots}
# stem count
ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = count_ha_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = count_ha_mean - count_ha_sd,
                    ymax = count_ha_mean + count_ha_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean stem count per cluster (n/ha) in year 30")

# volume/biomass
ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = volume_m3_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = volume_m3_mean - volume_m3_sd,
                    ymax = volume_m3_mean + volume_m3_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean volume per cluster (m³)/Biomass in year 30")

# DBH
ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = dbh_avg_cm_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = dbh_avg_cm_mean - dbh_avg_cm_sd,
                    ymax = dbh_avg_cm_mean + dbh_avg_cm_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean dbh per cluster (cm) in year 30")

# basal area
ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = basal_area_m2_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = basal_area_m2_mean - basal_area_m2_sd,
                    ymax = basal_area_m2_mean + basal_area_m2_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean basal area per cluster (m²) in year 30")

# cohort count
ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = cohort_count_ha_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = cohort_count_ha_mean - cohort_count_ha_sd,
                    ymax = cohort_count_ha_mean + cohort_count_ha_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean cohort count per cluster (n/ha) in year 30")
```

## 3.1 Interesting plots

Biomass development and biomass state in year 30. Mean volume per scenario and cluster (y = mean vol, x = cluster).

```{r interesting_plots}
# Development of biomass/volume over time
plot1 <- ggplot(w_means_clim, aes(x = year, y = volume_m3_mean, col = species)) +
  geom_line(aes(linetype = external_seed)) +
  geom_ribbon(aes(linetype = external_seed,
                  ymin = volume_m3_mean - volume_m3_sd,
                  ymax = volume_m3_mean + volume_m3_sd,
                  fill = species),
                  alpha = 0.05) +
  facet_wrap(~cluster, scales = "free") +
  theme_bw() +
  labs(title = "Development of mean volume(m³)/biomass per cluster")

# Volume/biomass state in year 30
plot2 <- ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = species, y = volume_m3_mean, fill = external_seed)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = volume_m3_mean - volume_m3_sd,
                    ymax = volume_m3_mean + volume_m3_sd)) +
  facet_wrap(~cluster, scales = "free_y", labeller = label_both) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  geom_text(aes(label = species), vjust = 0.25, check_overlap = TRUE) +
  labs(title = "Species mean volume per cluster (m³)/Biomass in year 30")

# Mean volume per scenario and cluster in year 30
plot3 <- ggplot(w_means_clim[w_means_clim$year == 30,],
       aes(x = cluster, y = volume_m3_mean, fill = species)) +
  geom_col(col = "black") +
  geom_text(aes(label = species),
            position = position_stack(vjust = 0.5),
            size = 4,
            check_overlap = TRUE) +
  facet_wrap(~external_seed, scales = "free", labeller = label_both) +
  theme_bw() +
  labs(title = "Mean volume/biomass per cluster in year 30")

plot1
plot2
plot3
```

## 3.2 Save plots

Safe the 3 plots as .png file.

```{r}
png("plots/plot1_year-vol.png", width = 1200, height = 1200)
plot1
dev.off()

png("plots/plot2_species-vol.png", width = 1200, height = 1200)
plot2
dev.off()

png("plots/plot3_cluster-vol.png", width = 1200, height = 1200)
plot3
dev.off()
```

# 4. Derive stand parameter

The development of the stand can also be predicted using 6 indicators, 4 of those should be calculated with iLand data (tree species, basal area, stem density, vertical class).

```{r calc_parameters}
w_means_sapl_clim$external_seed <- FALSE
w_means_sapl_clim[w_means_sapl_clim$ext_seed == "seed",]$external_seed <- TRUE
w_means_sapl_clim <- dplyr::select(w_means_sapl_clim, !ext_seed)

w_means_sapD_clim$external_seed <- FALSE
w_means_sapD_clim[w_means_sapD_clim$ext_seed == "seed",]$external_seed <- TRUE
w_means_sapD_clim <- dplyr::select(w_means_sapD_clim, !ext_seed)

# Gather data for every cluster
w_indic <- inner_join(w_means_clim, w_means_sapl_clim,
                      by = join_by(external_seed, cluster, year, species))

w_indic <- w_indic %>%
  dplyr::rename(count_ha_4_mean = count_ha_mean.x, # tree count/ha (>4m height)
                count_ha_4_sd = count_ha_sd.x,
                co_cnt_ha_mean = cohort_count_ha_mean.x, # number of cohorts (<4m height)
                co_cnt_ha_sd = cohort_count_ha_sd.x, # number of cohorts (<4m height)
                cnt_ha_13_mean = count_ha_mean.y, # number of individuals/ha (>1.3m height)
                cnt_ha_13_sd = count_ha_sd.y # number of individuals/ha (>1.3m height)
              )
w_indic <- inner_join(w_indic, w_means_sapD_clim,
                      by = join_by(external_seed, cluster, year, species))
w_indic %>%
  dplyr::rename(n_represented_coho_mean = n_represented_mean,
                n_represented_coho_sd = n_represented_sd,
                coho_dbh_mean = dbh_mean,
                coho_dbh_sd = dbh_sd,
                coho_age_mean = age_mean,
                coho_age_sd = age_sd)

```

Description of columns:

| Column name            | Description                                                                    |
|------------------------------------|------------------------------------|
| external_seed          | Wheather external seed was considered in the simulation                        |
| cluster                | To which of the 12 clusters this line belongs                                  |
| year                   | simulated year (0 is initial state, 30 final state)                            |
| species                | tree species the data relates to                                               |
| volume_m3_mean         | volume of species biomass (mean)                                               |
| volume_m3_sd           | volume of species biomass (standard deviation)                                 |
| dbh_avg_cm_mean        | average dbh of trees (\>4m height) (mean)                                      |
| dbh_avg_cm_sd          | average dbh of trees (\>4m height) (standard deviation)                        |
| basal_area_m2_mean     | total basal area at breast height (\>4m height) (mean)                         |
| basal_area_m2_sd       | total basal area at breast height (\>4m height) (standard deviation)           |
| count_ha_4_mean        | tree count (living, \>4m height) per ha (mean)                                 |
| count_ha_4_sd          | tree count (living, \>4m height) per ha (standard deviation)                   |
| co_cnt_ha_mean         | number of cohorts in the regeneration layer (\<4m) per ha (mean)               |
| co_cnt_ha_sd           | number of cohorts in the regeneration layer (\<4m) per ha (standard deviation) |
| cohort_basal_area_mean | basal area of saplings (\>1.3m and \<4m) (mean)                                |
| cohort_basal_area_sd   | basal area of saplings (\>1.3m and \<4m) (standard deviation)                  |
| cnt_ha_13_mean         | number of represented individuals per ha (\>1.3m height) (mean)                |
| cnt_ha_13_sd           | number of represented individuals per ha (\>1.3m height) (standard deviation)  |
| count_small_ha_mean    | number of represented individuals per ha (\<=1.3m height) (mean)               |
| count_small_ha_sd      | number of represented individuals per ha (\<=1.3m height) (standard deviation) |
| cohort_count_ha_mean.y | number of cohorts per ha (same as co_cnt_ha_mean)                              |
| cohort_count_ha_sd.y   | number of cohorts per ha (same as co_cnt_ha_sd)                                |
| height_avg_m_mean      | arithmetic average height of the cohorts (m) (mean)                            |
| height_avg_m_sd        | arithmetic average height of the cohorts (m) (sd)                              |
| age_avg_mean           | arithmetric average age of the sapling cohorts (years) (mean)                  |
| age_avg_sd             | arithmetric average age of the sapling cohorts (years) (standard deviaion)     |
| age_avg_sd             | arithmetric average age of the sapling cohorts (years) (standard deviaion)     |
| n_represented_mean     | number of trees that are represented by the cohort (Reineke function) (mean)   |
| n_represented_sd       | number of trees that are represented by the cohort (Reineke function) (sd)     |
| dbh_mean               | diameter of the cohort in cm (mean)                                            |
| dbh_sd                 | diameter of the cohort in cm (sd)                                              |
| height_mean            | height of the cohort in m (mean)                                               |
| height_sd              | height of the cohort in m (sd)                                                 |
| age_mean               | age of the cohort (mean)                                                       |
| age_sd                 | age of the cohort (sd)                                                         |

: column descriptions
