---
title: "Post-disturbance indicators for future stand development"
author: "Maria Potterf"
date: "2024-09-30"
output: 
  html_document:
    toc: true           # Enables the table of contents
    toc_depth: 3        # Sets the level of headers included in the TOC
    toc_float: true     # Makes the TOC float and always visible while scrolling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table) # Load data.table package
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)

library(networkD3)

# Set theme_classic2 as the default theme for all plots
theme_set(theme_classic2())
```

# Overview

In this overview, we focus on understanding of how how indicative the observation of the post-disturbance tree regeneration can be for future forest development. E.g., how will the structure and composition change in the future?

*Simulations*:

 - 12 simulated landscapes (3 climatic clusters, 4+5+3 structural clusters)
 - 3 climatic models = "ICHEC" "MPI"   "NCC"  
 - 4 climate scenarios = "HISTO" "RCP26" "RCP45" "RCP85"
 - 2 seed added/no = external seeds present or not? TRUE/FALSE
 - 5 repetitions
 - in total 1380 runs

landscape 3_2 does not have no seed scenario! (no regeneration present, so otherwise there would be no trees)




```{r read_data, include=FALSE}

# Simulated data:
# raw simulated data 
df_sim_class <- fread('outTable/fin_sim_data.csv')

# read simulated indicators
df_sim_indicators <- fread('outTable/df_simulated_indicators.csv')

# subset of field data: 12 landscapes
df_field_ind_sub <- fread('outTable/df_field_indicators_landscapes.csv')

# subset indicators to merge with sankey plot
df_field_ind_sub2 <- df_field_ind_sub %>% 
  dplyr::select(landscape, clim_cluster_spei3)

# compare 12 clusters: beginning (year 0) and end (year 25:30)
df_compare_end <- fread('outTable/compare_field_sim_12_lands_end.csv')

df_compare0 <- fread(      'outTable/compare_field_sim_12_lands_start.csv')


# read silhouette data - first tests!
sil_width_all <- fread("outTable/sil_width_test1.csv")


# read first sankey data
df_str_clustering_out <- fread('outTable/df_str_clustering_out_sankey.csv')


# Display the first few rows of the data
head(df_sim_class)
head(df_sim_indicators)

```


```{r functions, echo = F}
create_scatter_plot <- function(data, x_var, y_var, x_label, y_label, title) {
  # Get unique clusters and assign colors
  unique_clusters <- unique(data$landscape)
  
  # Create a color palette with enough colors for each cluster
  colors <- rainbow(length(unique_clusters))  # or any other palette function
  
  # Create the scatter plot using aes with .data[[var]]
  ggplot(data, aes(x = .data[[x_var]], y = .data[[y_var]], color = factor(landscape))) +
    geom_jitter(alpha = 0.2) +  # Apply alpha to jittered points
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(title = title, x = x_label, y = y_label) +
    theme_classic() +
    scale_color_manual(values = colors, 
                       guide = guide_legend(override.aes = list(size = 4, alpha = 1))) +
    facet_wrap(.~ext_seed)
}


```





```{r inspect_simulated, include=F}

unique(df_sim_class$landscape)  # 12, 3 climatic, 3-5 structural in each climate cluster
#[1] "1_1" "1_2" "1_3" "1_4" "2_1" "2_2" "2_3" "2_4" "2_5" "3_1" "3_2" "3_3"

unique(df_sim_class$species) # 24
#[1] "acps" "cabe" "potr" "saca" "quro" "acca" "frex" "ulgl" "lade" "tico" "piab" "fasy" "psme"
#[14] "soau" "pisy" "bepe" "abal" "alin" "algl" "soar" "coav" "acpl" "rops" "casa"


unique(df_sim_class$clim_model)
# [1] "ICHEC" "MPI"   "NCC"  

unique(df_sim_class$clim_scenario)
# [1] "HISTO" "RCP26" "RCP45" "RCP85"

# external seed: external seed present or not?? TRUE/FALSE
# years: 1-30
unique(df_sim_class$run_nr)  # 5 repetitions

```


# Interpret simulated data

##  climate-environment clusters:

-  clim_cluster_spei3  == 1 ~ "wet-warm-clay",  # Cluster 1: wet, cold, clay
-    clim_cluster_spei3  == 2 ~ "hot-dry-clay",  # Cluster 2: hot, dry, clay (more sand, less clay, more av.nitro than cluster 3)
-    clim_cluster_spei3  == 3 ~ "hot-dry-sand"    # Cluster 3: hot, dry, more sand

Each clim_env cluster has several structral clusters


```{r clim_clusters_diff, include=T, echo = F}

# explain climate and soil
p.tmp <- df_field_ind_sub %>% 
  ggplot(aes(x = clim_class,
             y = tmp)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))

p.prcp <- df_field_ind_sub %>% 
  ggplot(aes(x = clim_class,
             y = prcp)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))

p.spei <- df_field_ind_sub %>% 
  ggplot(aes(x = clim_class,
             y = spei3)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))

p.clay <- df_field_ind_sub %>% 
  ggplot(aes(x = clim_class,
             y = clay_extract)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5))


ggarrange(p.tmp, p.prcp, p.spei,p.clay, common.legend = T, ncol = 2, nrow = 2)


```

## Structural clusters:

- high density, high vertical layers, high richness, low rIVI

```{r structural_clusters_diff, include = T, echo = F}
# undetsrtand structural clusters
p1 <- df_field_ind_sub %>% 
  ggplot(aes(x = landscape,
             y = rIVI,
             fill = clim_class)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- df_field_ind_sub %>% 
  ggplot(aes(x = landscape,
             y = richness,
             fill = clim_class)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 <-df_field_ind_sub %>% 
  ggplot(aes(x = landscape,
             y = stem_density,
             fill = clim_class)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4<-df_field_ind_sub %>% 
  ggplot(aes(x = landscape,
             y = n_vertical,
             fill = clim_class)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggarrange(p1,p2,p3,p4, common.legend = T)


```

## Composite indicator 

order the landscapes by individual indicators. scale all indicators and add them together to order landscapes (from high structural and compositional  diversity to low structural and compositional diversity). 

Use it to simplify naming of the simulated landscapes

```{r check_landscape_composites, echo = F}
# par(mfrow = c(2, 2))
# plot(df_field_ind_sub$composite_score, 
#      df_field_ind_sub$stem_density)
# 
# plot(df_field_ind_sub$composite_score, 
#      df_field_ind_sub$n_vertical)
# 
# plot(df_field_ind_sub$composite_score, 
#      df_field_ind_sub$richness)
# 
# plot(df_field_ind_sub$composite_score, 
#      df_field_ind_sub$rIVI)


#par(mfrow = c(1, 1))
```


High composite score = high structural and compositional diversity

(my intension here was to order landscapes by the most diverse to the least diverse, but i have not implemented this all the way yet)


```{r get_landscapes_vs_composite}

df_field_ind_sub %>% 
  dplyr::select(composite_score,
               landscape) %>% 
  arrange(-composite_score)


```

# Plots

## 1:1 plots: 12 landscapes

check first on 12 landscapes: fit 1:1: field data vs simulated data in year 1

### Year 0: Beginning of simulation 

How simulated data in year 0 correspond to field data?

The data show teh same pattern within seed - no eeds scenarios, as there is not much influence of seeds in year 1


```{r compare_field_begin_simulation, echo = F}
# Create individual plots using the function
p0.1 <- create_scatter_plot(df_compare0, 
                          x_var = "rIVI_simul", 
                          y_var = "rIVI_field", 
                          x_label = "Simulated rIVI", 
                          y_label = "Field rIVI", 
                          title = "rIVI")

p0.2 <- create_scatter_plot(df_compare0, 
                          x_var = "richness_simul", 
                          y_var = "richness_field", 
                          x_label = "Simulated Richness", 
                          y_label = "Field Richness", 
                          title = "Richness")

p0.3 <- create_scatter_plot(df_compare0, 
                          x_var = "stem_density_simul", 
                          y_var = "stem_density_field", 
                          x_label = "Simulated Stem Density", 
                          y_label = "Field Stem Density", 
                          title = "Stem Density")

p0.4 <- create_scatter_plot(df_compare0, 
                          x_var = "n_vertical_simul", 
                          y_var = "n_vertical_field", 
                          x_label = "Simulated ", 
                          y_label = "Field ", 
                          title = "n Vertical")

# Arrange the plots in a 2x2 grid
ggarrange(p0.1, p0.2, p0.3, p0.4, ncol = 2, nrow = 2, common.legend = T, align = 'hv')




```

### Years 25:30 = End of simulation 

- including seeds lowers rIVI, increases richness, stem density and number of vertical layers. Red dashed line represents 1:1 scenarios (not different axes). 

Plots are jitter plots, therefre some values ay seems out of range.


```{r compare_field_end_simulation, echo = F}
# Create individual plots using the function
p1 <- create_scatter_plot(df_compare_end, 
                          x_var = "rIVI_simul", 
                          y_var = "rIVI_field", 
                          x_label = "Simulated rIVI", 
                          y_label = "Field rIVI", 
                          title = "rIVI: Simulated vs Field")

p2 <- create_scatter_plot(df_compare_end, 
                          x_var = "richness_simul", 
                          y_var = "richness_field", 
                          x_label = "Simulated Richness", 
                          y_label = "Field Richness", 
                          title = "Richness: Simulated vs Field")

p3 <- create_scatter_plot(df_compare_end, 
                          x_var = "stem_density_simul", 
                          y_var = "stem_density_field", 
                          x_label = "Simulated Stem Density", 
                          y_label = "Field Stem Density", 
                          title = "Stem Density: Simulated vs Field")

p4 <- create_scatter_plot(df_compare_end, 
                          x_var = "n_vertical_simul", 
                          y_var = "n_vertical_field", 
                          x_label = "Simulated n Vertical", 
                          y_label = "Field n Vertical", 
                          title = "n Vertical: Simulated vs Field")

# Arrange the plots in a 2x2 grid

ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE,
          align = 'hv')




```




## Temporal development of indicators: by landscapes

### Richness

```{r plot_richness, echo=FALSE, fig.width = 5, fig.height = 3}
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = richness,
                color = landscape,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom')+
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = richness, group = landscape, color = landscape), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')



```

### rIVI

```{r plot_rIVI, echo=FALSE,fig.width = 5, fig.height = 3}
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = rIVI,
                color = landscape,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom')+
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = rIVI, group = landscape, color = landscape), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')



```
### n_vertical
```{r plot_vertical, echo=FALSE, fig.width = 5, fig.height = 3}
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = n_vertical,
                color = landscape,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom') +
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = n_vertical, group = landscape, color = landscape), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')

```

### Stem density
```{r plot_stem_density, echo=FALSE,fig.width = 5, fig.height = 3}

df_sim_indicators %>% 
 # change in stem density
  ggplot() + 
  geom_line(aes(x = year,
                y = stem_density,
                color = landscape,
                group = unique_sim_run), alpha = 0.1) + 
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = stem_density, group = landscape, color = landscape), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(.~ext_seed)




```


# Temporal development: by advanced vs delayed regeneration


### Richness

```{r plot_richness2, echo=FALSE, fig.width = 5, fig.height = 3}

#head(df_sim_indicators)
###### Species richness ------------------------------------
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = richness,
                color = adv_delayed,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom')+
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = richness, group = adv_delayed, color = adv_delayed), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')



```

### rIVI

```{r plot_rIVI2, echo=FALSE,fig.width = 5, fig.height = 3}
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = rIVI,
                color = adv_delayed,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom')+
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = rIVI, group = adv_delayed, color = adv_delayed), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')



```
### n_vertical
```{r plot_vertical2, echo=FALSE, fig.width = 5, fig.height = 3}
###### NUmber of vertical classes ------------------------------------
df_sim_indicators %>% 
  ggplot() + 
  geom_line(aes(x = year,
                y = n_vertical,
                color = adv_delayed,
                group = unique_sim_run), alpha = 0.1) +
  theme(legend.position = 'bottom') +
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = n_vertical, group = adv_delayed, color = adv_delayed), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(. ~ext_seed) +
  theme(legend.position = 'right')

```

### Stem density
```{r plot_stem_density2, echo=FALSE,fig.width = 5, fig.height = 3}

df_sim_indicators %>% 
 # change in stem density
  ggplot() + 
  geom_line(aes(x = year,
                y = stem_density,
                color = adv_delayed,
                group = unique_sim_run), alpha = 0.1) + 
   # Add median line using stat_summary
  stat_summary(aes(x = year, y = stem_density, group = adv_delayed, color = adv_delayed), 
               fun = median, 
               geom = "line", 
               linewidth = 0.8) +
  facet_wrap(.~ext_seed)




```


# Structural cluster analysis

run using predictors: as no other variables were available! (disturbance severity, distance_edge, management_intensity)


data_subset_str  <- data_subset_str[, c("dominant_species", 
                                          "rIVI", 
                                          "richness",
                                          "stem_density",
                                          "n_vertical",
                                          "clim_cluster")]
                                          
Used k-means clustering coding the dominant species as dummy variable





```{r silhouette, echo = F, fig.width=3, fig.height = 6}
# Set up a 3-row layout for the plots
par(mfrow = c(3, 1))  # Arrange plots in a single column

# Calculate the global ylim to keep consistent scale across all plots
ylim_global <- c(0, max(sil_width_all, na.rm = TRUE))

# Plot for df1
plot(1:nrow(sil_width_all), sil_width_all$sil_width_df1, type = "b", 
     xlab = "Number of clusters", ylab = "Average Silhouette width", 
     main = "Silhouette Analysis for df1", ylim = ylim_global)

# Plot for df2
plot(1:nrow(sil_width_all), sil_width_all$sil_width_df2, type = "b", 
     xlab = "Number of clusters", ylab = "Average Silhouette width", 
     main = "Silhouette Analysis for df2", ylim = ylim_global)

# Plot for df3
plot(1:nrow(sil_width_all), sil_width_all$sil_width_df3, type = "b", 
     xlab = "Number of clusters", ylab = "Average Silhouette width", 
     main = "Silhouette Analysis for df3", ylim = ylim_global)


```

## Sankey plot: structural groups development

compare the structural clusters between the origin (field data) and future development (year 25:30). 

I selected only **single NO SEEDS scenario** here for visualization. Without seeds, the structurs are simplifying over time. 

```{r plot_sankey, echo=F}

# for individual groups

df_str_clustering_out2 <- df_str_clustering_out %>% 
  left_join(df_field_ind_sub2, by = join_by(landscape))

# Prepare a list to store plots
sankey_plots <- list()

# Loop through each unique clim_cluster
for (cluster in unique(df_str_clustering_out$clim_cluster)) {
  
  # Prepare data for the Sankey plot for the current clim_cluster
  sankey_data <- df_str_clustering_out %>%
    filter(clim_cluster == cluster) %>%
    group_by(str_cluster, str_cluster_end) %>%
    summarise(count = n(), .groups = 'drop')  # Count occurrences of each cluster pair
  
  # Create unique nodes
  nodes <- data.frame(name = unique(c(sankey_data$str_cluster, sankey_data$str_cluster_end)))
  
  # Create links data frame
  links <- sankey_data %>%
    mutate(source = match(str_cluster, nodes$name) - 1,  # Match the str_cluster to node indices
           target = match(str_cluster_end, nodes$name) - 1) %>%  # Match the str_cluster_end to node indices
    dplyr::select(source, target, value = count)  # Select the necessary columns
  
  # Create the Sankey plot for the current clim_cluster
  sankey_plot <- sankeyNetwork(Links = links, Nodes = nodes, Source = "source", Target = "target", 
                               Value = "value", NodeID = "name", units = "TWh", fontSize = 12, nodeWidth = 30)
  
  # Store the plot in the list
  sankey_plots[[as.character(cluster)]] <- sankey_plot
}



```

Clim Cluster 1
```{r plot_sankey1, echo=F, fig.width=5, fig.height=4}
# To view a specific plot, e.g., for the first clim_cluster
sankey_plots[[1]]  

```

Clim cluster 2
```{r plot_sankey2,fig.width=5, fig.height=4}
# To view a specific plot, e.g., for the first clim_cluster
sankey_plots[[2]]  

```

Clim cluster 3
```{r plot_sankey3,fig.width=5, fig.height=4}
# To view a specific plot, e.g., for the first clim_cluster
sankey_plots[[3]]  

```